\sc Estimating Evolutionary Parameters for Low Complexity Regions using an Approximate Bayesian Computation

### 2023-02-23 thesis_template/paper0904.tex
Feb 23, made some corrections from Brian in materials and methods

%\begin{figure}[h]
	\begin{enumerate}
		\item If now at $\theta$, propose a move to $\theta'$ according to a proposal distribution $q(\theta, \theta')$.
		\item Simulate a dataset, $D'$ using $\theta'$.
		\item If $D' ~ D$ proceed to step 4; else, output $\theta$ and return to 1.
		\item Calculate the Hastings Ratio.
		\item Accept, and output, the new $\theta'$ with probability h. Else return to, and output, $\theta$, Go back to 1.
	\end{enumerate}
	%\label{fig:algorithm1}
	%\caption{Hello}
%\end{figure}

A custom \texttt{C++} script was written to iterate through the algorithm for a desired number of simulations. The Normal distribution was used to control how new parameter values were proposed \hl{SHOULD I MENTION THE MEAN AND STDEV OF THE NORMAL HERE? WHAT IF I WANT TO TEST MANY DIFFERENT KINDS}. We simulated a dataset under the newly proposed parameter value, which consisted of a randomly generated protein sequence. The simulated protein was compared to a protein of known low complexity called SRP40, which is found in the model organism \sclong. The protein sequence for SRP40 was downloaded from the NCBI database. To quantitatively compare similarities between simulated and observed protein sequences, vectors of summary statistics (characteristics that describe the sequences) were created for each sequence. The euclidean distance between the two vectors was calculated and compared to a threshold value to determine if the newly proposed parameter value could be accepted.

\hl{NEED HELP:} In order to create a posterior distribution, parameter values need to be accepted with a certain probability. In the algorithm proposed by \citet{marjoram2003markov}, the Hastings Ratio is calculated to determine this acceptance probability. Currently having issues implementing the hastings algorithm and struggling to understand why I can not use it or figure it out. What are other alternatives in this case for calculating acceptance probabilities?

@@

Two parameters were estimated in this study, mutation rate and insertion/deletion (indel) rate. Mutation rate referred to the rate at which a single amino acid in a protein sequence changed into a different amino acid. Indel rate referred to the rate at which an amino acid was deleted or inserted from a protein sequence. For the purpose of this study, to determine if the length of a repetitive region played a role in insertions, any amino acid that was inserted into the simulated protein sequence was the same unit as the previous amino acid in the sequence \hl{(Should I explain why here, not too sure, going to talk more about it in the next part).}

Summary statistics were utilized to capture important information about simulated and observed protein sequences in order to assess how similar the sequences were. Three summary statistics were used which included, the length of the protein sequence, the number of LCRs in the sequence, and the average entropy of the LCRs. To identify LCRs and their corresponding entropies, the \texttt{Seg} algorithm was implemented \citep{wootton1993statistics}. The following \texttt{Seg} parameters were utilized to search for LCRs in proteins; a window length of 15, a trigger segment complexity of 1.9, and an extension segment complexity of 2.2. We selected these due to previous research which demonstrated that these parameter values would better detect regions of low complexity in eukaryotes which are typically longer and contain more repetitive repeats \citep{huntley2000evolution, huntley2002simple}. \hl{In the 2002 paper, the 2000 paper is cited for this fact so I was not sure which to cite, put both for now.}

@@

\hl{Need Help:} We currently do not know how well these summary statistics explain the data. Weighting summary statistics could be beneficial, but determining how we weight each one appropriately is challenging at the moment. There are potential weighting techniques, such as one method used by \citet{hamilton2005bayesian} in which summary statistics are regressed on retained parameters, allowing for a weight to be calculated for each summary statistic and subsequently used when calculating the Euclidean Distance. This method is very difficult to implement and I am struggling. Maybe other alternatives for this weighting?

@@

Exponential deviates were stored in two vectors, one for deviates generated using mutation rate, and the other for deviates generated using the indel rate. We then identified a single deviate with the lowest value (based on both vectors), which represented the residue that mutated quickest. Depending on which vector the lowest deviate came from, we either altered the corresponding amino acid or inserted/deleted a repetitive amino acid.

We ran the simulation ten times for each newly proposed parameter value

@@

\section{Appendix C++ simulation (so far)}
\lstinputlisting[language=c++]{mutations_2.cpp}

>>>>

%\begin{figure}[h]
	\begin{enumerate}
		\item If now at $\theta$, propose a move to $\theta'$ according to a proposal distribution $q(\theta, \theta')$.
		\item Simulate a dataset, $D'$ using $\theta'$.
		\item If $D' \approx D$ proceed to step 4; else, output $\theta$ and return to 1.
		\item Calculate the Hastings Ratio.
		\item Accept, and output, the new $\theta'$ with probability h. Else return to, and output, $\theta$, Go back to 1.
	\end{enumerate}
	%\label{fig:algorithm1}
	%\caption{Hello}
%\end{figure}

A custom \texttt{C++} script was written to iterate through the algorithm for a desired number of simulations. The Normal distribution was used to control how new parameter values were proposed. We simulated a dataset under the newly proposed parameter value, which consisted of a randomly generated protein sequence. The simulated protein was compared to a protein of known low complexity called SRP40, which is found in the model organism \sclong. The protein sequence for SRP40 was obtained from the NCBI database. To quantitatively compare similarities between simulated and observed protein sequences, vectors of summary statistics (characteristics that describe the sequences) were created for each sequence. \hl{The Euclidean distance between the two vectors was calculated and compared to a threshold value to determine if the newly proposed parameter value could be accepted.}

@@

Two parameters were estimated in this study, mutation rate and insertion/deletion (indel) rate. Mutation rate referred to the rate at which a single amino acid in a protein sequence changed into a different amino acid. Indel rate referred to the rate at which an amino acid was deleted or inserted from a protein sequence. For the purpose of this study, to determine if the length of a repetitive region played a role in insertions, any amino acid that was inserted into the simulated protein sequence was the same unit as the previous amino acid in the sequence.

Summary statistics were utilized to capture important information about simulated and observed protein sequences in order to assess how similar the sequences were. Three summary statistics were used which included, the length of the protein sequence, the number of LCRs in the sequence, and the average entropy of the LCRs. To identify LCRs and their corresponding entropies, the \texttt{Seg} algorithm was implemented \citep{wootton1993statistics}. The following \texttt{Seg} parameters were utilized to search for LCRs in proteins; a window length of 15, a trigger segment complexity of 1.9, and an extension segment complexity of 2.2. We selected these due to previous research which demonstrated that these parameter values would better detect regions of low complexity in eukaryotes which are typically longer and contain more repetitive repeats \citep{huntley2000evolution}.

@@

@@

Exponential deviates were stored in two vectors, one for deviates generated using mutation rate, and the other for deviates generated using the indel rate. We then identified a single deviate with the lowest value (based on both vectors), which represented the residue that mutated quickest. Depending on which vector the lowest deviate came from, we either altered the corresponding amino acid or inserted/deleted a repetitive amino acid. Upon mutating, deleting, or inserting an amino acid, we scanned the sequence to see if the mutation altered the landscape (inturrupted a repeat, created a repeat), and subsequently generated new deviates for the amino acids that were affected.

We ran the simulation ten times for each newly proposed parameter value and took the average of all ten vectors of summary statistics prior to calculating the distance. On each simulation, we mutated the protein sequence for 2000 generations before obtaining summary statistics. We plan on testing the program in the future with various numbers of iterations.

@@

\section{Appendix C++ simulation (so far)}
\lstinputlisting[language=c++]{../mutations_2_BG_vecs.cpp}
